Inference for multivariate peaks over threshold models

### 2021-10-27 results.tex
overleaf updates

\section{Results}

\subsection{Simulation}
We have introduced this projected gamma distribution existing on $\mathcal{S}_{\infty}^{p-1}$.  As a
  means of evaluating how well we can recover a distribution under the proposed model, we have
  generated a series of datasets on $\mathcal{S}_{\infty}^{d-1}$, for $d$ ranging between 3 and 20.
  We generated each dataset from a mixture of projected Gammas, with the number of mixture components
  ranging between 3 and 12.  The generation procedure is detailed in Algorithm~\ref{algo:simulated}.
  \begin{algorithm}[htb]
    \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}{
      Generate $n_{\text{mix}} \times 20$ shape Parameters $\alpha$\\
      Generate $n_{\text{mix}} \times 20$ Rate Parameters $\beta$\\
      Generate $500$ Mixture Component Identifiers $\delta$\\
      \For{$i$ in $1,\ldots,500$}{
        Generate ${\bf X}_i \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{il}\mid\alpha_{\delta_i,l},\beta_{\delta_i, l}\right)$
        }
      \For{$n_{\text{col}}$ in $[3,6,12,20]$}{
        Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
        }
    }
    \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
  \end{algorithm}
  The choice of generation routine for $\alpha = \alpha_0 + \alpha_1$ where
  $\alpha_0 \sim \text{Unif}(\alpha_0\mid 0,4)$, $\alpha_1\sim \text{Gamma}(\alpha_1\mid 1,1)$.
  The choice of generation routine for $\beta$ is $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.
  This choice was made to ensure that $\beta$ does not approach 0 in the generation routine.

\begin{figure}[ht]
  \caption{Simulation Energy Score versus Column Count, by Mixture Component Count.\label{fig:simes}}
  \centering
  \includegraphics[width=0.8\textwidth]{./images/simulation_es}
\end{figure}

In Figure~\ref{fig:simes}, we fit DP Mixtures of the various models presented and plot the resulting
  energy score against the number of columns, grouped by the number of mixture components used in
  dataset generation.  The \emph{Generative} model refers to a dataset generated
  in the same manner, from the same parameters as was used to create the original dataset.  Against
  this generative dataset, we compare DP Mixtures of projected gamma, projected restricted gamma, and
  projected restricted gamma with a multivariate log-normal prior.  We observe that the projected
  restricted Gamma model dominates in most situations, but the difference between projected restricted
  Gamma and the other gamma-based models tends to shrink as the number of columns increases.
  Alternatively, that difference appears to grow as the number of mixture components increases.

\begin{figure}[ht]
  \caption{Simulation Posterior Predictive Loss versus Column Count, by Mixture Component Count.\label{fig:simppl}}
  \centering
  \includegraphics[width=0.8\textwidth]{./images/simulation_ppl}
\end{figure}

@@

\begin{figure}[ht]
  \caption{Estimated $k$NN-KL Divergence metric on simulated data, averaged between 1 and 10 nearest
    neighbors, against column count, for selected models.  Graphs are facetted according to the
    number of mixture components in the generating distribution.  The \emph{generative} model
    represents another replicate of the underlying distribution, so that should be regarded as
    \emph{truth}.\label{fig:simkld}}
  \includegraphics[width = 5.5in, height = 3.75in]{./images/simulation_knn_kld}
\end{figure}

in Figure~\ref{fig:simkld} we see the application of the $k$NN-based K.L. divergence metric detailed
  in Equation~\ref{eqn:knnkld} to the simulated data sets.  For display, this metric has been averaged
  between $1$ and $10$ neighbors.  In general, the K.L. divergence metric is a log-ratio
  of densities, integrated over the entire space.  In this fashion, we should interpret that model
  is \emph{best} that stays closest to 0, as that indicates the two densities are equivalent.  The
  equation also includes a scaling factor for the size of the data sets, as all else being equal, a
  larger data set will decrease the distance between any point and its neighbors.  We can regard the
  \emph{generative} model as another replicate data set from the generating distribution---another
  view of truth.  As its distribution should be almost identical, we would assume that the KL divergence
  metric from this model will come closest to 0.  It does not.  Knowing this to be the case, we should
  instead regard the generative model as truth, and treat as best the model which comes closest to the
  \emph{generative} model.  In this regard, we observe that the projected gamma model comes closest,
  with the projected restricted gamma model following behind.  The probit-normal model quickly diverges
  as the number of dimensions increases.  This is not unexpected, as we saw similar behavior in the
  energy score and posterior predictive loss criteria.

\subsection{Integrated Vapor Transport}
We use data from the integrated water vapor transport (\emph{IVT}) model \citep{guan2015}
  of atmospheric rivers as a means of testing these models.  An atmospheric river is a meteorological
  event of local concentration of water vapor in the atmosphere that moves with wind patterns.
  Understanding extremal dependence in atmospheric rivers, for instance, grants hydrological engineers
  a greater understanding of how and where to build ground water capacity to handle precipitation.
  California suffers persistent, extended droughts---if ground water transportation and holding capacity
  (canals, reservoirs) are not built to manage extreme precipitation without adequate understanding of
  the extremal dependence, then we may not install adequate capacity and thus experience flooding.
  Alternatively, if we install too much capacity everywhere, then we are wasting money that could be
  better spent elsewhere.

@@

\begin{algorithm}
  \KwResult{$\bm{ r} : r_i \sim \text{Pareto}(1)$, $\bm{ v} \in \mathcal{S}_{\infty}^{d-1}$}
  \For{$\ell = 1,\ldots,d$}{
    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\chi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
    }
  \For{$i = 1,\ldots,n$}{
    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ V}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
    }
  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$ for all $r_i\in \bm{r}$.\\
  \If{declustering}{
    \For{$i = 1,\ldots,n$}{
      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
    }
  }
 \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
   portion is relevant if data is correlated in time.\label{algo:processing}}
\end{algorithm}

@@

\begin{table}[b]
  \centering
  \caption{Model comparison metrics: Posterior Predictive Loss and Energy Score criteria from fitted
    models against the IVT data.  All presented models are DP mixtures; the \emph{Model} field
    identifies the kernel distribution.  For both criteria, lower is better.
  \label{tab:dev}}
  \include{table_dev}
\end{table}

In Table~\ref{tab:dev} we see the same model preference towards the projected restricted gamma models
  that we saw in Figures~\ref{fig:simes},\ref{fig:simppl}.  The unrestricted gamma model is penalized much
  more strongly on real data than we saw with our simulation, performing comparably to the probit-normal
  model.  We also see the restricted gamma model with the log-normal prior performing significantly
  better than the restricted gamma model with the gamma prior on the higher dimensional data, a reversal
  of what we saw on the low dimensional data and indeed from what we saw in the simulation study.

That said, there is much we can glean from this table.  The probit-normal model, developed on a
  mapping from $\mathcal{S}_{2}^{d-1}$ to $(-\infty,\infty)^{d-1}$ via probit transformation of
  spherical coordinates was not competitive with the gamma models on real data, similar to what we
  saw in the simulation study. We also established finite mixture models for all presented
  distributions, but after tuning the number of mixture components, their model performance was,
  and could only be \emph{as good} as the DP mixture.  In the interest of parsimony, we only present
  the DP models as representative of mixture distributions.

\begin{figure}[ht]
  \centering
  \caption{KL Divergence Curves calculated through the KNN-KL Metric, evaluated between the
  empirical dataset and posterior predictive datasets of various models.  The top row corresponds
  to the 8-dimensional data, while the bottom row the 47.  All presented models use a DP prior,
  the nominal model being the kernel distribution.\label{fig:knnkl}}
  \includegraphics[width = 5.5in, height = 2.75in]{./images/knn_kld}
\end{figure}

Interpreting the KL divergence curves in Figure~\ref{fig:knnkl} is more difficult.  As we saw in
  the simulated case in Figure~\ref{fig:simkld}, the general logic of \emph{best} being that which
  is closest to zero is not necessarily in use here.  We interpret a positive value in the KL
  divergence metric as indicating that, for a given point in the given distribution, we must
  travel \emph{farther} on the target distribution to reach the $k$th neighbor as compared to
  the given distribution.  In this regard, we should regard \emph{large} positive values as
  indicating a particularly poor fit.  By this metric, the projected gamma and probit-normal
  models do not appear to scale well as they performed very poorly on the high dimensional data.
  For the projected restricted gamma model, the log-normal prior appears to scale better than the
  gamma prior.  This is in keeping with what we observed in Table~\ref{tab:dev}, but is a
  different result than what we observed in the simulation study, where we observed that the
  log-normal prior only tended towards the gamma prior, never performing better.

>>>>

\section{Data illustrations\label{sec:results}}

\subsection{Simulations\label{subsec:simulated}}
We have introduced this projected gamma distribution existing on $\mathcal{S}_{\infty}^{p-1}$.  As a
  means of evaluating how well we can recover a distribution under the proposed model, we have
  generated a series of datasets on $\mathcal{S}_{\infty}^{d-1}$, for $d$ ranging between 3 and 20.
  We generated each dataset from a mixture of projected Gammas, with the number of mixture components
  ranging between 3 and 12.  The generation procedure is detailed in Algorithm~\ref{algo:simulated}.
  \begin{algorithm}[ht]
    \caption{Simulated Angular Dataset Generation Routine\label{algo:simulated}}
    \For{$n_{\text{mix}}$ in $[3, 6, 9, 12]$}{
      Generate $n_{\text{mix}} \times 20$ shape Parameters $\alpha$\\
      Generate $n_{\text{mix}} \times 20$ Rate Parameters $\beta$\\
      Generate $500$ Mixture Component Identifiers $\delta$\\
      \For{$i$ in $1,\ldots,500$}{
        Generate ${\bf X}_i \sim \prod_{\ell = 1}^d\text{Ga}\left(X_{il}\mid\alpha_{\delta_i,l},\beta_{\delta_i, l}\right)$
        }
      \For{$n_{\text{col}}$ in $[3,6,12,20]$}{
        Project columns 1 to $n_{\text{col}}$ of ${\bf X}$ onto 
        $\mathcal{S}_{\infty}^{n_{\text{col}} - 1}$ and save.
        }
    }
  \end{algorithm}
  The choice of generation routine for $\alpha = \alpha_0 + \alpha_1$ where
  $\alpha_0 \sim \text{Unif}(\alpha_0\mid 0,4)$, $\alpha_1\sim \text{Gamma}(\alpha_1\mid 1,1)$.
  The choice of generation routine for $\beta$ is $\beta\sim\text{Unif}(\beta\mid 0.25, 2.5)$.
  This choice was made to ensure that $\beta$ does not approach 0 in the generation routine.

\begin{figure}[ht]
  \caption{Simulation Energy Score versus Column Count, by Mixture Component Count. Lower is better.\label{fig:simes}}
  \centering
  \includegraphics[width=0.8\textwidth]{./images/simulation_es}
\end{figure}

In Figure~\ref{fig:simes}, we fit DP Mixtures of the various models presented and plot the resulting
  energy score against the number of columns, grouped by the number of mixture components used in
  dataset generation.  The \emph{Gen} model refers to a dataset generated
  in the same manner, from the same parameters as was used to create the original dataset.  Against
  this generative dataset, we compare DP Mixtures of projected gamma, projected restricted gamma, and
  projected restricted gamma with a multivariate log-normal prior.  We observe that the projected
  restricted Gamma model dominates in most situations, but the difference between projected restricted
  Gamma and the other gamma-based models tends to shrink as the number of columns increases.
  Alternatively, that difference appears to grow as the number of mixture components increases.

\begin{figure}[ht]
  \caption{Simulation posterior predictive loss versus column count, by mixture component count.\label{fig:simppl}}
  \centering
  \includegraphics[width=0.8\textwidth]{./images/simulation_ppl}
\end{figure}

@@

\subsection{Integrated Vapor Transport\label{subsec:ivt}}
The \emph{integrated vapor transport} (IVT) tracks the total water volume in a column of air over a given
  area \citep{ralph2017}.  IVT is the means by which we track the formation and path of atmospheric 
  rivers---elongated areas of high local concentration of water vapor in the atmosphere that moves with
  wind patterns.  Extreme behavior in IVT can lead to extreme precipitation events, causing flooding; 
  but in California, those same events are responsible for the bulk of precipitation that inland 
  California recieves---water necessary for agriculture and household consumption.  Thus it becomes 
  necessary to manage precipitation to prevent inundation and support agriculture.  To this end, 
  understanding the dependence structure of extreme events in the IVT is useful---for instance, for 
  hydrological engineers in planning the capacity of groundwater systems to capture precipitation before 
  wasteful and destructive flooding occurs.  Our data tracks daily average values for the IVT, dividing 
  the coast of California into grid cells, with two spatial resolutions of 8 and 47 cells, respectively 
  \citep{guan2015}.

@@

\begin{algorithm}
  \KwResult{$\bm{ r},\bm{v} : r_i \sim \text{Pareto}(1)$, $\bm{ v}_i \in \mathcal{S}_{\infty}^{d-1}$}
  \For{$\ell = 1,\ldots,d$}{
    Set $b_{t,\ell} = \hat{F}_{\ell}^{-1}\left(1 - \frac{1}{t}\right)$.\\
    With $\bm{ x}_{\ell} > b_{t,\ell}$, fit $a_{\ell}$, $\chi_{\ell}$ via MLE according to generalized Pareto likelihood.\\
    }
  \For{$i = 1,\ldots,n$}{
    Define $z_{i,\ell} = \left(1 + \xi_{\ell}\frac{x_{i,\ell} - b_{t,\ell}}{a_{\ell}}\right)_{+}^{1/\xi_{\ell}}$\\
    Define $r_i = \pnorm{\bm{ z}_i}{\infty}$, $\bm{ v}_i = \frac{\bm{ z}_i}{\pnorm{\bm{ z}_i}{\infty}}$\\
    }
  Subset $\bm{ r},\bm{ v}$ such that $r_i \geq 1$ for all $r_i\in \bm{r}$.\\
  \If{declustering}{
    \For{$i = 1,\ldots,n$}{
      If $r_i \geq 1$ and $r_{i-1} \geq 1$, drop the lesser (and associated $v_i$) from data set.\\
    }
  }
 \caption{Data preprocessing to isolate and transform data exhibiting extreme behavior.  $r_i$
   represents the radial component, and $\bm{v}_i$ the angular component.  The declustering
   portion is relevant if data is correlated in time.\label{algo:processing}}
\end{algorithm}

@@

\begin{table}[ht]
  \centering
  \caption{Model comparison metrics: Posterior Predictive Loss and Energy Score criteria from fitted
    models against the IVT data.  Lower is better.
  \label{tab:dev}}
  \include{./tables/table_dev}
\end{table}

In Table~\ref{tab:dev} we see the same model preference towards the projected restricted gamma models
  that we saw in Figures~\ref{fig:simes}.  The unrestricted gamma model is penalized much
  more strongly on real data than we saw with our simulation, performing comparably to the probit-normal
  model.  We also see the restricted gamma model with the log-normal prior performing significantly
  better than the restricted gamma model with the gamma prior on the higher dimensional data, a reversal
  of what we saw on the low dimensional data and indeed from what we saw in the simulation study.
